---
title: "ECS230 Hw3"
author:
- Eric Kalosa-Kenyon
date: "`r format(Sys.time(), '%d %B %Y')`"

abstract: In this assignment, I time matrix multiplications to assess processor performance in basic linear algebra operations.

output:
  html_document:
    toc: yes
    smart: false
---

# Matrix multiplication
In this section, I report the results of the timing subroutines developed in the
source code included in the last section of this report.

## Repeated multiplications
```{r, echo=F}
library("knitr")
read_chunk("../src/hw3.R")
```

```{r SETUP, echo=F}
```

```{r PR1, echo=F}
```

## Loop rearrangement
Using column-major ordering (i.e. $i + n*j$) for array indexing into the $i$th
row and $j$th column. The following experiment times different orderings of the
embedded for loops used to calculate the matrix product. A typical ordering
may look like the following:

```
sum = 0
for(i in 1:n){
    for(j in 1:n){
        for(k in 1:n){
             sum += A[i + k*n] * B[k + i*n]
        }
        C[i + n*j] = sum
        sum = 0
    }
}
```

The preceeding code snippet reflects the following matrix product definition:
$$
A_{n\times n} B_{n\times n} = C_{n\times n} = (c_{i,j})_{1\le i, j\le n} =
(\sum_{k=1}^n a_{i, k} b_{k, j}) =
(\langle \vec{a_i^\top}, \vec{b_j} \rangle )_{1\le i,j\le n}
$$

```{r PR2, echo=F}
```

The loop orderings are as follows, using column-major indexing:
1. jik
2. jki
3. ijk
4. ikj
5. kji
6. kij

## Using BLAS dgemm()
```{r PR3, echo=F}
```

## Comments on performance
There were no attempts to make use of block structures in this multiplication.
The result of this negligance is that once the size of the matrix exceeds the
size of the processor's L1 cache, the speed of computation is massively reduced.

The number of flops theoretically required to perform a basic matrix
multiplication is $2n^3$. The $2$ is due to the addition and multiplication in
$s = s + a_{i,k}b_{k,j}$ and the $n^3$ is due to the three embedded for loops.

# Source code
```{r engine='bash', comment=''}
# cat ../src/mmult.c
# cat ../src/all_looptypes_mmult.c
# cat ../src/dgemm_mmult.c
```
