---
title: "ECS230 Hw5"
author:
- Eric Kalosa-Kenyon
date: "`r format(Sys.time(), '%d %B %Y')`"

abstract: In this assignment, I implement parts of PageRank.

output:
  html_document:
    toc: yes
    smart: false
---

# Brief proof of a property of eigenvalues
Let $A$ be a real $n\times n$ matrix and $\lambda$ an eigenvalue of $A$. Here,
I show that $\lambda$ is an eigenvalue of $A^\top$ as well.

Notice that the eigenvalues of $A$ are the roots of its characteristic
polynomial $f(\lambda) = det(A-\lambda I)$. Similarly, let
$f^\top(\lambda) = det(A^\top-\lambda I)$ be the characteristic polynomial of
the transposed matrix $A^\top$. This proof boils down to showing that
$f(\lambda) = f^\top(\lambda)$ as follows:

1. $f(\lambda) = det(A-\lambda I)$
2. $det(A-\lambda I) = det((A-\lambda I)^\top)$
3. $det((A-\lambda I)^\top) = det(A^\top-\lambda I^\top)$
4. $det(A^\top-\lambda I^\top) = det(A^\top - \lambda I)$
5. $det(A^\top - \lambda I) = f^\top(\lambda) \quad \blacksquare$



```{r, echo=F}
library("knitr")
read_chunk("../src/hw5.R")
```

```{r READ, echo=F, message=F}
```

```{r GRAPH, echo=F}
```

# Appendix
The following sections contain useful supplementary information.

## Data

```{r engine='bash', comment='bryan_leise_2006_fig1.dat:', echo=F}
cat ../data/bryan_leise_2006_fig1.dat
```

```{r engine='bash', comment='ex1_add_pg_5.dat:', echo=F}
cat ../data/ex1_add_pg_5.dat
```

```{r engine='bash', comment='lin_chain_len_4.dat:', echo=F}
cat ../data/lin_chain_len_4.dat
```

```{r engine='bash', comment='test_A.dat:', echo=F}
cat ../data/test_A.dat
```

## Source code
The following is the source used for the creation of this report. The code
handed in on the CSIF computers is slightly different from the following source.
The hand-in code conforms to the specifications of the assignment, wheras the
following source is formatted to work with the project's directory structure. No
major algorithmic changes were made.

```{r engine='bash', comment='Main source:', echo=F}
cat ../src/power_method.c
```

```{r engine='bash', comment='', echo=F}
cat ../src/hw5.R
```
