---
title: "ECS230 Hw5"
author:
- Eric Kalosa-Kenyon
date: "`r format(Sys.time(), '%d %B %Y')`"

abstract: In this assignment, I implement parts of PageRank.

output:
  html_document:
    toc: yes
    smart: false
---

# Brief proof of a property of eigenvalues
Let $A$ be a real $n\times n$ matrix and $\lambda$ an eigenvalue of $A$. Here,
I show that $\lambda$ is an eigenvalue of $A^\top$ as well.

Notice that the eigenvalues of $A$ are the roots of its characteristic
polynomial $f(\lambda) = det(A-\lambda I)$. Similarly, let
$f^\top(\lambda) = det(A^\top-\lambda I)$ be the characteristic polynomial of
the transposed matrix $A^\top$. This proof boils down to showing that
$f(\lambda) = f^\top(\lambda)$ as follows:

1. $f(\lambda) = det(A-\lambda I)$
2. $det(A-\lambda I) = det((A-\lambda I)^\top)$
3. $det((A-\lambda I)^\top) = det(A^\top-\lambda I^\top)$
4. $det(A^\top-\lambda I^\top) = det(A^\top - \lambda I)$
5. $det(A^\top - \lambda I) = f^\top(\lambda) \quad \blacksquare$

Warrant for each of the steps in the proof above follow:

The first and final steps of the proof are the respective definitions of
$f(\lambda)$ and $f^\top(\lambda)$. The second step can be seen by expanding
$A-\lambda I$ into its $QR$ decomposition and using determinant product rules.
The third and fourth steps are obvious.

# Power method
To solve for the dominant eigenvalue of a matrix $A$, the power method can be
used. In short, the method is as follows:
$$b_{k+1} = \frac{Ab_k}{||Ab_k||}$$
This method and several extensions are implemented in the source code included
at the end of this report.

## Reproducing a simple result
```{r echo=F, message=F}
library("knitr")
read_chunk("../src/plot_graphs.R")
```

```{r read_library, echo=F, message=F}
```

```{r graph1, echo=F, message=F}
```

Using the graph presented in figure 1 of Bryan and Leise (2006), the program
produes the following vector as the dominant L1-normalized eigenvector:
$$v_1 =
\left(
\begin{matrix}
0.387097 \\
0.129032 \\
0.290323 \\
0.193548
\end{matrix}
\right)
$$
This solution is produced using the "bryan_leise_2006_fig1.dat" input file and
corresponds to the ranking $A, C, D, B$.

## A PageRank manipulation strategy

```{r graph2, echo=F, message=F}
```

Using the strategy described in exercise 1 of Bryan and Leise where the owners
of page $C$ make a new page $E$ and link both together, the dominant eigenvector
becomes
$$v_2 =
\left(
\begin{matrix}
0.244898 \\
0.081633 \\
0.367347 \\
0.122449 \\
0.183673
\end{matrix}
\right)
$$
This new dominant eigenvector $v_2$ corresponds to the ranking
$C, A, E, D, B$. Notice that, using this strategy, the owners of page $3$ were
able to manipulate the ranks.

## A chain graph and the shifted power method

```{r graph3, echo=F, message=F}
```

Using a chain graph of length $n=4$ and an initial vector
$$b_0 =
\left(
\begin{matrix}
1.000000 \\
1.000000 \\
1.000000 \\
1.000000
\end{matrix}
\right)
$$

my program converges to the dominant eigenvector
$$v_3 =
\left(
\begin{matrix}
0.166667 \\
0.333333 \\
0.333333 \\
0.166667
\end{matrix}
\right)
$$

corresponding to the ranking $\{2, 3\}, \{1, 4\}$ as expected. However, due to
the fact that the spectrum of the induced matrix column-stochastic link matrix
$A$ is
$$
spec(A) = spec(
\left(
\begin{matrix}
0 & 1/2 & 0 & 0 \\
1 & 0 & 1/2 & 0 \\
0 & 1/2 & 0 & 1 \\
0 & 0 & 1/2 & 0
\end{matrix}
\right)
) = \{1, \frac{1}{2}, \frac{-1}{2}, -1\}
$$

yielding two dominant eigenvectors corresponding to the eigenvalues $1$ and
$-1$, the power method fails to converge for impropperly selected initial
vectors. For example, when using
$$b_0 =
\left(
\begin{matrix}
1.00000 \\
2.00000 \\
3.00000 \\
4.00000
\end{matrix}
\right)
$$

the result oscilates between
$$
\left(
\begin{matrix}
0.20000 \\
0.26666 \\
0.40000 \\
0.13333
\end{matrix}
\right)
\textrm{ and }
\left(
\begin{matrix}
0.372104 \\
0.496139 \\
0.744208 \\
0.248069
\end{matrix}
\right)
$$

To solve this problem of non-convergence, I implement the shift method. In
short, this replaces the matrix $A$ with $B=A+cI$ for some constant $c$. Note
that the eigenvectors are identical for $A$ and $B$: if $v$ is an eigenvector of
$A$, then $Bv=(A+cI)v=Av+cv=(\lambda+c)v$. Here, noting that the spectrum is
bounded by $-1$ and $1$, I use $c=\frac{1}{4}$ to force the desired dominant
eigenvector to have an eigenvalue of $1.25$ and bring the other eigenvectors
away from dominance.

Upon implementation, the shifted power method converges for all tested initial
vectors.

# Appendix

## Data

Following are the files used as inputs for the power_method program and their
respective outputs. Each input
represents a link matrix and follows the specification for .dat files given in
the assignment handout:
1. The first line is a single integer $n$ stating the number of nodes in the
   graph.
2. Each subsequent line $j$ starts with an integer $n_j$ stating the number of
   outgoing links from page $j$.
3. On line $j$, after the initial $n_j$, each subsequent integer $i$ points to
   an external page $i$ to which page $j$ links.


bryan_leise_2006_fig1.dat:
```{r engine='bash', comment='input', echo=F}
cat ../data/bryan_leise_2006_fig1.dat
```

```{r engine='bash', comment='output', echo=F}
cat ../data/bryan_leise_2006_fig1.out
```

ex1_add_pg_5.dat:
```{r engine='bash', comment='input', echo=F}
cat ../data/ex1_add_pg_5.dat
```

```{r engine='bash', comment='output', echo=F}
cat ../data/ex1_add_pg_5.out
```

lin_chain_len_4.dat:
```{r engine='bash', comment='input', echo=F}
cat ../data/lin_chain_len_4.dat
```

using the initial vector $b_0=[1,1,1,1]$,
```{r engine='bash', comment='output', echo=F}
cat ../data/lin_chain_len_4.out
```

using the initial vector $b_0=[1,2,3,4]$,
```{r engine='bash', comment='output', echo=F}
cat ../data/lin_chain_len_4.out2
```

using the initial vector $b_0=[1,2,3,4]$ and a shift of $c=\frac{1}{4}$,
```{r engine='bash', comment='output', echo=F}
cat ../data/lin_chain_len_4.out3
```

## Source code
The following is the source used for the creation of this report. The code
handed in on the CSIF computers is slightly different from the following source.
The hand-in code conforms to the specifications of the assignment, wheras the
following source is formatted to work with the project's directory structure. No
major algorithmic changes were made.

```{r engine='bash', comment='', echo=F}
cat ../src/power_method.c
```

```{r engine='bash', comment='', echo=F}
cat ../src/plot_graphs.R
```
